Best Practices in Latent Class Analysis using Free Open Source Software

### 2022-10-28 manuscript.Rmd
edits from Mauricio

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LCA"

author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"

affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"

author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands

abstract: |
  Latent class analysis is a family of techniques for identifying groups in data based
  on a parametric model. Its examples include mixture models, latent profile
  analysis, latent class analysis with ordinal indicators, and latent class
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on
  the `OpenMx` R-package.

keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "8206"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no

class             : "man"
output            : papaja::apa6_pdf

---

@@

Latent class analysis (LCA) is an umbrella term 
that refers to a number of techniques for estimating unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile analysis.
@vermunt_jk_latent_2004 defined LCA more generally as virtually any statistical model 
where “some of the parameters […] differ across unobserved subgroups”.

@@

Latent class analysis is a group of methods for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of LCA can be understood in different ways.
A mixture model assumes that the study population
is composed of $K$ subpopulations or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate model is a normal distribution, 
which can be described with two parameters: the mean and the variance.
Commonly, the same model is estimated across all classes, 
but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class, 
and the probability that an individual belongs to each class.

@@

A LCA is like confirmatory factor analysis, 
except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that 
factor analysis can be considered as a way to group observed *variables* into latent constructs, 
and LCA groups *individuals* into classes.
In line with this distinction, LCA is sometimes referred to as a "person-centered" technique, 
and factor analysis as a "variable-centered" technique [@nylund-gibson_ten_2018].

@@

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm.
In line with this perspective, LCA is sometimes described as "model-based
clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast "model-based" clustering refers to the fact that LCA 
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

@@

A common type of LCA is the *finite Gaussian mixture model*; 
a univariate analysis where the observed distribution of a single variable 
is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
The parameters of a finite Gaussian mixture model are 
the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivariate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances 
between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or set to be constrained across classes.

The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that, after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for instance when the cohesion between indicators is expected to differ between classes.
As an example, a LPA of ocean plastic particles 
found two classes of particles based on length and width:
a class of smaller particles with a high correlation between length and width, 
meaning that these particles were approximately round or square in shape, 
and a class of larger particles with a low correlation between length and width, 
meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense, 
because the smaller particles have been ground down to a more uniform shape by the elements.
<!-- DA: what is the source for plastic particles? -->

It is also possible to estimate a mixture model based on latent indicators.
This means that, within each class, 
one or more continuous latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated 
based on these continuous latent variables.
A common application of this approach is in longitudinal research, 
where the indicators reflect one construct assessed at different time points.
Examples of this approach include *growth mixture models* (GMM) 
and *latent class growth analyses* (LCGA).
These techniques estimate a latent growth model
to describe individual trajectories over time.
The growth mixture model is a latent class model 
where the parameters that indicate class membership 
are the intercepts and variances, 
and typically covariances of the latent growth variables, 
e.g., a latent intercept and slope.
This technique assumes that individuals within a class can have heterogenous trajectories.
If the variance of the growth parameters is fixed to zero, 
it is known as a latent class growth analysis.
This latter approach assumes that all individuals within a class 
share the same identical trajectory,
and that any variance in the indicators not explained by the class-specific latent trajectories 
is due to residual error variance.

The term latent class analysis originally referred to cases 
where the observed indicators were categorical.
Nowadays, it is more commonly used as an umbrella term.
To prevent ambiguity, the special case 
where indicators are of binary or ordinal measurement level 
might be described as *latent class analysis with ordinal indicators*.
Latent class models with ordinal indicators are parameterized differently from mixture models.
One common parameterization assumes that 
each categorical variable reflects an underlying standard normal distribution.
The parameters are "thresholds" that correspond to quantiles of a standard normal distribution 
(with $N(\mu = 0, \sigma = 1)$).
These thresholds are estimated based on 
the proportion of individuals in each of the response categories of the indicator variable.
For example, a binary indicator has a single threshold that distinguishes the two response categories.
If responses are distributed 50/50, then the corresponding threshold would be  $t_1 = 0.00$.
If the responses are distributed 60/40, then the resulting threshold would be $t_1 = 0.25$.

@@

Another use case is unsupervised learning:
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid 
when several clinical indicators can be used to distinguish between 
a fixed number of physical [@baughman_mixture_2006] or mental [@wu_abuse_2011] health problems.
The example of shoe size is a rudimentary illustration of this type of application.

@@

@@

<!-- CJ: (on the next paragraph)  -->
<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->
<!-- That is not what this paragraph is about.  -->
<!-- This paragraph is about motivations for using LCA. -->
<!-- One motivation is if your data violates multivariate normality -->
<!-- because violating multivariate normality means that you have multiple groups -->
<!-- and you can use LCA to extract those. That is the only assumption  -->
<!-- we have to discuss here, we just need to explain it. -->
Finally, LCA can be used to deal with data which violate certain assumptions.
As discussed in the shoe size example, LCA can deal with violations of normality.
In fact, LCA assumes the population distribution is a non-normal mixture 
of $K$ normal distributions,
and it can discover the value of $K$, i.e. generate a $K$-class solution.

@@

Examining observed data is essential for any analysis
as it may reveal patterns and violations of assumptions 
that had not been considered prior to data collection.
Special attention should be paid 
to level of measurement of the indicators.
Finite Gaussian mixture models (including LPA) 
are only suitable for continuous variables.
Indicators with an ordinal level of measurement 
are likely to violate the assumption of 
within-class normal distributions of mixture models
[see @vermunt_k-means_2011].
Personal experience consulting on LCA methods and 
moderating the `tidyLPA` Google group
suggest that the misapplication of mixture models 
to ordinal (e.g., Likert-type) indicators 
is the most common source of user error.
Whereas it has been argued that some parametric methods are robust 
when scales with 7+ indicators are treated as continuous 
[e.g., @norman_likert_2010],
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables 
can be treated as a *mixture* of multiple normal distributions.
The problem becomes egregious when 
the number of classes estimated equals or exceeds the number of categories; 
in this case, each class-specific mean could describe a single response category, 
and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that 
an assumption of mixture models is that 
observed indicators are normally distributed [@spurk_latent_2020].
This is incorrect.
When the number of classes is greater than one, 
mixture models assume that the observed indicators 
are a mixture of multiple (multivariate) normal distributions.
In our shoe size example, it can be seen that the population distribution
is comprised of two normal distributions.
When examined visually, the population distribution is evidently bimodal. 
The Shapiro-Wilk normality test ($W = 0.971$, $p < 0.05$ ) 
rejects the null hypothesis that 
the sample comes from a normally distributed population.
Yet, this is a prototypical example of a mixed population distribution 
where LCA can discover latent groups.
If the population distribution were instead normal, 
there would be no classes to extract 
as the whole population would belong to a single class.

@@

Previous work has emphasized the importance of examining 
the pattern of missing data and 
reporting how missingness was handled 
[@van_de_schoot_grolts-checklist_2017].
Three types of missingness have been distinguished
in the literature [@rubin_inference_1976]:
Missing completely at random (MCAR), 
which means that missingness is random;
missing at random (MAR), 
which means that missingness is contingent on the *observed* data
(and can thus be accounted for);
and finally missing not at random (MNAR), 
which means that missingness is related to unobserved factors.
It is possible to conduct a so-called "MCAR" test, 
for example the non-parametric MCAR test [@jamshidian_tests_2010].
But note that the name "MCAR test" is somewhat misleading,
as the null-hypothesis of this test is that the data are not MAR,
and a significant test statistic indicates that 
missingness is related to the observed data (MAR).
A non-significant test statistic does not distinguish between MCAR or MNAR.
As Little's classic MCAR test relies on 
the comparison of variances across groups 
with different patterns of missing data, 
it assumes normality [@little_test_1988].
This assumption is tenuous in the context of LCA.
A non-parametric MCAR test, 
as provided by Jamshidian and Jalal, 
may be more suitable [@jamshidian_tests_2010].
Unfortunately, this test was removed from the central R-repository CRAN 
due to lack of maintenance.
For this tutorial, I have re-implemented it in the `mice` package 
as `mice::mcar()`,
with a fast backend in C++ and new printing and plotting methods.

While we concur that investigating missingness is due diligence,
it is important to emphasize that 
missingness is adequately handled by default in many software packages for LCA, 
such as Mplus, and `OpenMx` which is the backend of `tidySEM`.
These packages use Full Information Maximum Likelihood (FIML) estimation,
which makes use of all available information without imputing missing values.
FIML is a best-practice solution for handling missing data; 
on par with multiple imputation [@lee_comparison_2021].
FIML estimation assumes that missingness is either MCAR or MAR.
Thus, one would typically proceed with FIML, 
regardless of the outcome of an MCAR test.
Although FIML does not, by default, handle missingness in exogenous variables
-- all indicator variables in LCA are endogenous, 
so this is not a concern.

@@

Conversely, the free open source program OpenMx is very flexible,
but not very user-friendly. 
We directly address this limitation using the `tidySEM` R-package.
New functionality in `tidySEM` seeks to lower the threshold 
for latent class analysis using `OpenMx`.
It adheres to best practices in estimation and reporting, 
as described in this paper.
The user interface is simple, 
making use of the model syntax of the widely used `lavaan` R-package.
This syntax offers a human-readable way to specify latent variable models.
Minor enhancements are made to simplify the specification of LCA.

@@

A recent paper suggested maximum likelihood with robust standard errors 
should be used when the observed indicators are not normally distributed [@spurk_latent_2020]. 
This statement is incorrect, and may lead readers to believe that 
they must use commercial software, 
as robust maximum likelihood is currently only implemented 
in Mplus and latentGOLD.
As explained before, 
mixture modeling assumes that observed data are 
a mixture of (multivariate) normal distributions; 
thus, the observed indicators will likely not be normally distributed.
<!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->

@@

From a sequence of models, the final class solution is chosen 
based on both theoretical and statistical criteria.
Theory should drive the selection of indicator variables, 
inform the expectations and reflect on the findings.
In addition to this, there are several statistical criteria 
to consider in model selection.
These include but are not limited to likelihood ratio tests, 
information criteria, 
and the Bayes factor [@weller_latent_2020].

Relative model fit can be examined using the likelihood ratio test. 
This is only appropriate when the two models we wish to compare are nested.
The likelihood ratio test statistic is computed as 
the difference in maximum log likelihoods of the two models,
with the test degrees of freedom being 
the difference in the degrees of freedom of the two compared models.
The test statistic follows the $\chi^2$ distribution,
and we want it to be non-significant
in order to give support to the simpler model. 
The likelihood ratio test can only compare two nested models at a time [@lanza_introduction_2003].

@@

Fit indices typically used for determining the optimal number of classes include
the Akaike Information Criterion (AIC) 
and the Bayesian Information Criterion (BIC).
Both information criteria are based on the -2*log likelihood 
(which is lower for better fitting models),
and add a penalty for the number of parameters (thus incentivizing simpler models).
This helps balance model fit and model complexity.
The lower the value of an information criterion, 
the better the overall fit of the model.
The BIC applies a stronger penalty for model complexity 
that scales logarithmically with the sample size.
The literature suggests the BIC may be 
the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018].
Both the AIC and the BIC are available in the `tidySEM` output.

@@

LCA studies commonly report -2*log likelihood of the final class solution.
This is a basic fit measure used to compute most information criteria.
However, since log likelihood is not penalized for model complexity,
it will continuously fall with the addition of more classes.

@@

Three important classification diagnostics provided by `tidySEM` are: 
(1) *the minimum* and *maximum percentage of the sample assigned to a particular class*,
(2) *the range of the posterior class probabilities by most likely class membership*, 
and (3) *entropy*. 
All three are based on posterior class probabilities.

@@

`$individual` is the individual posterior probability matrix, 
with dimensions n (number of cases in the data) x k (number of classes).
Individual class probabilities are often useful for researchers
who wish to do follow up analyses.

@@

Class names should be chosen to accurately reflect group membership. 
Overly simplified and generalized class names may prove misleading 
to both audiences and researches alike
leading to what is known as a naming fallacy [@weller_latent_2020].

@@

TOP-guidelines emphasise the use of comprehensive citation 
(including referencing the software used in the analysis),
as well as code and data sharing wherever possible [@nosek_promoting_2015].
@van_lissa_worcs_2020 suggest sharing synthetic data in case 
the original data cannot be shared, 
and provide functions to generate such synthetic data.
Ideally, the entire research project is made reproducible 
so that others may download it and reproduce it with just one click; 
for guidance, see @van_lissa_worcs_2020.

>>>>

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LCA"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis is a family of techniques for identifying groups in data based
  on a parametric model. Its examples include mixture models, latent class
  analysis (with continuous and/or ordinal indicators), and latent class (mixture)
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on
  the `OpenMx` R-package.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

@@

Latent class analysis (LCA) is an umbrella term that refers to a number
of techniques for estimating unobserved group membership based on a
parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile
analysis. @vermunt_jk_latent_2004 defined LCA more generally as
virtually any statistical model where "some of the parameters [...]
differ across unobserved subgroups".

@@

Latent class analysis is a group of methods for estimating unobserved
groups based on a parametric model of observed indicators of group
membership. The concept of LCA can be understood in different ways. A
mixture model assumes that the study population is composed of $K$
subpopulations or classes. It further assumes that the observed data are
a mixture of data generated by class-specific models. The simplest
univariate model is a normal distribution, which can be described with
two parameters: the mean and the variance. Commonly, the same model is
estimated across all classes, but with different parameters for each
class (i.e., class-specific means and variances). Mixture modeling then
estimates both the parameters for each class, and the probability that
an individual belongs to each class. This way, describing the whole
sample with a single mean and variance is a special case where we assume
the sample is homogeneous in function of these parameters.

@@

LCA is like confirmatory factor analysis, except that the continuous
latent variable is substituted with a categorical latent variable. Both
assume that the indicators are related, and these relations can be
represented by the underlying latent variable. One difference between
the two techniques is that factor analysis can be considered as a way to
group observed *variables* into latent constructs, and LCA groups
*individuals* into classes.While factor analysis seeks to describe the
relations between variables, LCA seeks to describe each unobserved group
and the differences between them. In line with this distinction, LCA
(mixture models) is referred to as a "person-centered" technique, and
factor analysis as a "variable-centered" technique
[@nylund-gibson_ten_2018; @masyn_latent_2013].

@@

In factor analysis, the latent variable is continuous, so the relation
between indicators is define by the communalities between them, and the
latent variable represents the error corrected approximation of the
sample. While in LCA, the relation between indicators is define by how
well they separate the classes, as a multinomial regression between
indicators and latent variable [@masyn_latent_2013]

When the focus is on each individual's estimated class membership, LCA
can be thought of as a type of clustering algorithm. In line with this
perspective, LCA is sometimes described as "model-based clustering"
[@hennig_handbook_2015; @scrucca_mclust_2016]. Many clustering
algorithms apply some recursive splitting algorithm to the data. By
contrast "model-based" clustering refers to the fact that LCA estimates
cluster membership based on a parametric model. Specifically, the
posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data
under given the class-specific model. For these differences, LCA is a clustering method, that corrects for (random) measurement error, while clustering algorithms do not correct for. This is something LCA has in common with factor analysis, as both model an underlying latent variable, which accounts for the communalities between indicators and correcting for measurement error. 0

@@

A common type of LCA is the *finite Gaussian mixture model*; a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be
estimated freely, or set to be constrained across classes [@collins_latent_2010].

The technique known as *latent profile analysis (LPA)* is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for, the covariances/correlations between indicators are
assumed to be zero. This can be conceived of as a restricted mixture
model with covariances fixed to zero. Stating that the covariances/correlations are accounted by the latent variable. In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes [@collins_latent_2010]. As an example, a LPA of ocean
plastic particles found two classes of particles based on length and
width: a class of smaller particles with a high correlation between
length and width, meaning that these particles were approximately round
or square in shape, and a class of larger particles with a low
correlation between length and width, meaning that these particles were
heterogenous in shape. From a theoretical perspective, this makes sense,
because the smaller particles have been ground down to a more uniform
shape by the elements.
<!-- DA: what is the source for plastic particles? -->

It is also possible to estimate a mixture model based on latent
indicators. This means that, within each class, one or more continuous
latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated based on these
continuous latent variables. A common application of this approach is in
longitudinal research, where the indicators reflect one construct
assessed at different time points. Examples of this approach include
*growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
These techniques estimate a latent growth model to describe grouped
trajectories over time. The growth mixture model is a latent class model
where the parameters that indicate class membership are the intercepts
and variances, and typically covariances of the latent growth variables,
e.g., a latent intercept and slope. This technique assumes that
individuals within a class can have heterogenous trajectories. If the
variance of the growth parameters is fixed to zero, it is known as a
latent class growth analysis. This latter approach assumes that all
individuals within a class share the same identical trajectory, and that
any variance in the indicators not explained by the class-specific
latent trajectories is due to residual error variance.
<!-- MGV: some reference for this description -->

The term latent class analysis originally referred to cases where the
observed indicators were categorical [@collins_latent_2010]. 
Nowadays, it is more commonly used
as an umbrella term. To prevent ambiguity, the special case where
indicators are of binary or ordinal measurement level might be described
as *latent class analysis with ordinal indicators*. Latent class models
with ordinal indicators are parameterized differently from mixture
models. One common parameterization assumes that each categorical
variable reflects an underlying standard normal distribution. The
parameters are "thresholds" that correspond to quantiles of a standard
normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
are estimated based on the proportion of individuals in each of the
response categories of the indicator variable. For example, a binary
indicator has a single threshold that distinguishes the two response
categories. If responses are distributed 50/50, then the corresponding
threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
then the resulting threshold would be $t_1 = 0.25$.

@@

Another use case is unsupervised learning: when the goal is to restore
unobserved class membership based on observed indicators, or to classify
individuals. For example, a mixture model can be used as a diagnostic
aid when several clinical indicators can be used to distinguish between
a fixed number of physical [@baughman_mixture_2006], mental
[@wu_abuse_2011] health problems, or health outcomes after recovering from cancer [@hardy_exploring_2018]. The example of shoe size is a
rudimentary illustration of this type of application.

@@

These mixture models can also be useful to evaluate quality of measurement across instrument. For example, employment status was collected across register and survey data, and there is disagreement between them. You could think that register data is without error, but it has been shown that is victim of measurement error as well. In this case mixture models was used to evaluate the underlying factor of "error corrected employment status", by using two indicators of the same question and identify random and systamtic measurement error in either indicator [@pavlopoulos_measuring_2015].

@@

<!-- CJ: (on the next paragraph)  -->

<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->

<!-- That is not what this paragraph is about.  -->

<!-- This paragraph is about motivations for using LCA. -->

<!-- One motivation is if your data violates multivariate normality -->

<!-- because violating multivariate normality means that you have multiple groups -->

<!-- and you can use LCA to extract those. That is the only assumption  -->

<!-- we have to discuss here, we just need to explain it. -->

Finally, LCA can be used to deal with data which violate certain
assumptions. As discussed in the shoe size example, LCA can deal with
violations of normality. In fact, LCA assumes the population
distribution is a non-normal mixture of $K$ normal distributions, and it
can discover the value of $K$, i.e. generate a $K$-class solution.

LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous of categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question.

@@

Examining observed data is essential for any analysis as it may reveal
patterns and violations of assumptions that had not been considered
prior to data collection. Special attention should be paid to level of
measurement of the indicators. Finite Gaussian mixture models (including
LPA) are only suitable for continuous variables. Indicators with an
ordinal level of measurement are likely to violate the assumption of
within-class normal distributions of mixture models [see
@vermunt_k-means_2011]. Personal experience consulting on LCA methods
and moderating the `tidyLPA` Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
indicators are treated as continuous [e.g., @norman_likert_2010] or 
with 5+ categories [@rhemtulla_when_2012], this
certainly does not imply that all methods are. It is certainly unlikely
that such ordinal variables can be treated as a *mixture* of multiple
normal distributions. The problem becomes egregious when the number of
classes estimated equals or exceeds the number of categories; in this
case, each class-specific mean could describe a single response
category, and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for Gaussian mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that an assumption of mixture
models is that observed indicators are normally distributed
[@spurk_latent_2020]. This is incorrect. When the number of classes is
greater than one, Gaussian mixture models assume that the observed indicators are
a mixture of multiple (multivariate) normal distributions. In our shoe
size example, it can be seen that the population distribution is
comprised of two normal distributions. When examined visually, the
population distribution is evidently bimodal. The Shapiro-Wilk normality
test ($W = 0.971$, $p < 0.05$ ) rejects the null hypothesis that the
sample comes from a normally distributed population. Yet, this is a
prototypical example of a mixed population distribution where LCA can
discover latent groups. If the population distribution were instead
normal, there would be no classes to extract as the whole population
would belong to a single (homogeneous) class.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
[@van_de_schoot_grolts-checklist_2017]. Three types of missingness have
been distinguished in the literature [@rubin_inference_1976; @enders_applied_2010]: Missing
completely at random (MCAR), which means that missingness is random and unrelated to any observed or unobserved variables;
missing at random (MAR), which means that missingness is contingent on
the *observed* data (and can thus be accounted for); and finally missing
not at random (MNAR), which means that missingness is related to
unobserved factors. It is possible to conduct a so-called "MCAR" test,
for example the non-parametric MCAR test [@jamshidian_tests_2010]. But
note that the name "MCAR test" is somewhat misleading, as the
null-hypothesis of this test is that the data are not MAR, and a
significant test statistic indicates that missingness is related to the
observed data (MAR). A non-significant test statistic does not
distinguish between MCAR or MNAR. As Little's classic MCAR test relies
on the comparison of variances across groups with different patterns of
missing data, it assumes normality [@little_test_1988]. This assumption
is tenuous in the context of LCA. A non-parametric MCAR test, as
provided by Jamshidian and Jalal, may be more suitable
[@jamshidian_tests_2010]. Unfortunately, this test was removed from the
central R-repository CRAN due to lack of maintenance. For this tutorial,
I have re-implemented it in the `mice` package as `mice::mcar()`, with a
fast backend in C++ and new printing and plotting methods.

While we concur that investigating missingness is due diligence, it is
important to emphasize that missingness is adequately handled by default
in many software packages for LCA, such as Mplus, Latet GOLD, and `OpenMx` which is
the backend of `tidySEM`. These packages use Full Information Maximum
Likelihood (FIML) estimation, which makes use of all available
information without imputing missing values. FIML is a best-practice
solution for handling missing data; on par with multiple imputation when the estimation model includes the same variables as the imputation model
[@lee_comparison_2021]. FIML estimation assumes that missingness is
either MCAR or MAR. Thus, one would typically proceed with FIML,
regardless of the outcome of an MCAR test. Although FIML does not, by
default, handle missingness in exogenous variables -- all indicator
variables in LCA are endogenous, so this is not a concern.
<!-- are you sure Mplus accounts for missing data automatically when the indicators are categorical? I know that is not the case with CFA/SEM models -->

@@

Conversely, the free open source program OpenMx [@openmx] is very flexible, but
not very user-friendly. We directly address this limitation using the
`tidySEM` R-package. New functionality in `tidySEM` seeks to lower the
threshold for latent class analysis using `OpenMx`. It adheres to best
practices in estimation and reporting, as described in this paper. The
user interface is simple, making use of the model syntax of the widely
used `lavaan` R-package [@lavaan]. This syntax offers a human-readable way to
specify latent variable models. Minor enhancements are made to simplify
the specification of LCA.

@@

A recent paper suggested maximum likelihood with robust standard errors
should be used when the observed indicators are not normally distributed
[@spurk_latent_2020]. This statement is incorrect, and may lead readers
to believe that they must use commercial software, as robust maximum
likelihood is currently only implemented in Mplus and LatentGOLD. As
explained before, mixture modeling assumes that observed data are a
mixture of (multivariate) normal distributions; thus, the observed
indicators will likely not be normally distributed.
<!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->

@@

From a sequence of models, the final class solution is chosen based on
both theoretical and statistical criteria. Theory should drive the
selection of indicator variables, inform the expectations and reflect on
the findings. In addition to this, there are several statistical
criteria to consider in model selection. These include but are not
limited to likelihood ratio tests, information criteria, and the Bayes
factor [@weller_latent_2020; @masyn_latent_2013].

Relative model fit can be examined using the likelihood ratio test. This
is only appropriate when the two models we wish to compare are nested.
The likelihood ratio test statistic is computed as the difference in
maximum log-likelihoods of the two models, with the test degrees of
freedom being the difference in the degrees of freedom of the two
compared models. The test statistic follows the $\chi^2$ distribution,
and we want it to be non-significant in order to give support to the
simpler model. The likelihood ratio test can only compare two nested
models at a time [@lanza_introduction_2003].

@@

Fit indices typically used for determining the optimal number of classes
include the Akaike Information Criterion (AIC) and the Bayesian
Information Criterion (BIC). Both information criteria are based on the
-2\*log-likelihood (which is lower for better fitting models), and add a
penalty for the number of parameters (thus incentivizing simpler
models). This helps balance model fit and model complexity. The lower
the value of an information criterion, the better the overall fit of the
model.

The general objective of the Information Criteria (IC) is to evaluate the model's out-of-sample predictive accuracy, so adjusting for over fitting. Fit measures like $R^2$ evaluate the in-sample predictive accuracy, meaning that they evaluate the models ability to predict the observed outcomes based on the same data that was use to build up the model. These metrics are positively bias, meaning that will present better model fit than it has in reality. While IC corrects for this positive bias by evaluating the models accuracy approximating the out-of-sample predictive accuracy, meaning that it is the ability to predict the outcome for observations that are not part of the training model. Ideally we would estimate the model a lot of time with less observations and predict their scores, but IC's approximate this by their different penalty metrics. For example $R^2$ will increase even if an added predictors are unnecessary, while IC's will show worst fit when a predictor (complexity) is unnecessary [@mcelreath_statistical_2020]

The BIC applies a stronger penalty for model complexity that
scales logarithmically with the sample size. The literature suggests the
BIC may be the most appropriate information criterion to use for model
comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. Both the AIC and the BIC are
available in the `tidySEM` output.

@@

A disadvantage of the IC's is that we do not have standard errors for them, so we only have the absolute values without a measure of their variability. So, the difference between models IC can be very small. Still indicating that the model with the lower value is "better", but if this difference is  very small can considered them "functionally equal", and you should take into consideration the interpretability of the model.

LCA studies commonly report -2\*log likelihood of the final class
solution. This is a basic fit measure used to compute most information
criteria. However, since log likelihood is not penalized for model
complexity, it will continuously fall with the addition of more classes.

@@

Three important classification diagnostics provided by `tidySEM` are:
(1) the *minimum* and *maximum* percentage of the sample assigned to a
particular *class*, (2) the *range of the posterior class probabilities* by most likely class membership, and (3) *entropy*. All three are based
on posterior class probabilities.
<!-- MGV: I will work on adding average posterior class probability (AvePP) as well-->

@@

`$individual` is the individual posterior probability matrix, with
dimensions n (number of cases in the data) x k (number of classes). Additionally it includes the `predicted` class in function of the highest predicted probability.
Individual class probabilities and/or predicted class are often useful for researchers who wish to do follow up analyses.

@@

Class names should be chosen to accurately reflect group membership, as theoretically relevant names according to the interpretation of the class solutions.
Overly simplified and generalized class names may prove misleading to
both audiences and researches alike leading to what is known as a naming
fallacy [@weller_latent_2020].

@@

TOP-guidelines emphasize the use of comprehensive citation (including
referencing the software used in the analysis), as well as code and data
sharing wherever possible [@nosek_promoting_2015]. @van_lissa_worcs_2020
suggest sharing synthetic data in case the original data cannot be
shared, and provide functions to generate such synthetic data. Ideally,
the entire research project is made reproducible so that others may
download it and reproduce it with just one click; for guidance, see
@van_lissa_worcs_2020.

As the open science movement is gaining momentum, researchers are
becoming increasingly aware how important it is that analyses can be
reproduced and audited. In line with open science principles, one of the
suggested reporting standards relates to reproducible code. In this
context, it is important to note that user-friendly methods for
estimating LCA models have predominantly been available in commercial
software packages (e.g., *Mplus* and *Latent GOLD*). A potential
downside of commercial software is that it restricts the ability to
reproduce analyses to license holders, and prevents auditing research
because the underlying source code is proprietary. To overcome these
limitations, the present paper introduces new user-friendly functions in
the `tidySEM` R-package that can be used to estimate a wide range of LCA
models using the free, open-source R-package `OpenMx` as the backend. The reporting
guidelines described in this paper are adopted in `tidySEM` by default.
The `tidySEM` R-package thus makes advanced mixture modeling based on
best practices widely accessible, and facilitates the adoption of the
estimation and reporting guidelines described in this paper.
